---
title: "Programming Language Project"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, comment = "---", message = FALSE, collapse = TRUE)
```

This is an assignment for my programming languages class. My goal is to make an accessibly-written tutorial for people who are new to programming. I'm very new to github and R, so please be patient as we learn together.

Thank you!

## PLP I - Getting Started

### A Brief History
R is a programming language and coding environment designed for statistical computation. It is also a foundation that centers itself around the goal of public good by providing access to free, open-source statistical coding software. It was developed in the mid-nineties and version 1.0 was officially released in 2000. It sits at number 18 on the TIOBE index (at the time of writing) and is one of the most utilized languages (if not THE most) in the field of data analytics. Its name is a reference to its predecessor, S, and doubles as a reference to its founders, Robert Gentleman and Ross Ihaka, whose names share the letter as their first initials.

### Usages of R
R is open-source and free on principle, making it a very accessible and popular tool in statistical education and data analytics. Perhaps its most popular and powerful trait is its community-driven nature; its utility is constantly expanded by user-created "packages" that are easy to install and use for an extremely adaptable experience. That's not to say that it only finds usage in independent academics and researchers though -- R is reported to be used for the collection and processing of user data for giants like Amazon, Meta, and Google. It's also used by the FDA and National Weather Service to create models and visualize data for the public.

### R Resources (R-esources?)
[The R Foundation](https://www.r-project.org/foundation/), 
[RStudio (Currently being renovated)](https://www.rstudio.com), and 
[Book Recommendations By RStudio](https://www.rstudio.com/resources/books/)

### Downloading and Using R
R can be downloaded from [r-project.org](https://cran.r-project.org/mirrors.html), and its signature desktop IDE (Integrated Development Environment -- essentially, the place where you write and run code), RStudio, can be downloaded from [rstudio.com](https://posit.co/download/rstudio-desktop/), or used as an online application called [RStudio Cloud](https://posit.cloud/plans). It is also supported by VSCode (tutorial by VSCode [here](https://code.visualstudio.com/docs/languages/r)). **All of this software is available for free!**

### Hello, World!
Time for the time-honored programming tradition of the "Hello, World!" code! Lucky for us, this one's simple. With your IDE of choice, use a print statement:

```{r}
print("Hello World!")
```

### Comments in R
To include human-friendly text within code to clarify things for non-computer audiences, you can "comment" your code. This text is ignored by the computer when your program is run.

Unfortunately, R does not currently support multi-line commenting, but single-line commenting is allowed and a keyboard command makes commenting out larger sections easy. Use ```#``` to comment out a single line like so:

```{r}
#Anything I write here is for humans to read
print("This is a sample line of code") #And over here

# And all these lines
# are "commented out"
# as well
```

For multi-line comments (for example, a block of code you want to temporarily remove from your program), highlight the section and use the command ```ctrl``` + ```shift``` + ```c```

## PLP 2 - Data Types and Naming Conventions

### Naming conventions
In R, "snake_case," "camelCase," and "dot.case" are all valid naming conventions. Typically, snake case and camel case are preferred to avoid confusion with object-oriented programming languages. Additionally, when naming functions, the first letter is capitalized to distinguish them from variables.

Variables should not start with an underscore, a dot, or a number.

Reserve words in R include TRUE, FALSE, NULL, NA, NaN, Inf, if, else, repeat, while, for, in, function, next, break, NA_character_, NA_integer_, NA_real_, NA_complex_, and ellipses (...).

### Variable typing

R is an implicitly typed language, meaning that a variable's type does not have to be immediately declared. It is dynamically typed, meaning that the program only "knows" the variable type once the program is run. And given its nature as a statistics-based language, it is weakly typed, meaning that it will generally accommodate the programmer's wishes when asked to execute operations between data types (ie. a float and an integer).

### Types of variables
- Strings: Essentially, text variables. Written in quotes.
- Numerics: Variables that are numbers, including integers (which are exclusively whole numbers, written with a capital L immediately following the value) and floats (variables that are numbers that either have decimal points or may have them later, no special notation needed)
- Lists: A list of variables (of any type) that are stored together. Written in parentheses and separated by commas
- Boolean values: Variables used for logical operations: TRUE and FALSE
- Hashes: Similar to a list, but each item has a corresponding item attached to it. This data type is often called a "dictionary" in other languages due to its "key/definition" structure. This requires the "hash" package, which means you need to run the command ```library(hash)``` before creating a hash.

Examples:

```{r}
## Strings

string <- "string"
print(string)
```

```{r}
## Numerics (Integer)

numericint <- 5L
print(numericint)
```

```{r}
## Numerics (Float)

numericfloat <- 5.5
print(numericfloat)
```

```{r}
## Interactions between types of numerics

print(numericint + numericfloat)
print(numericint * numericfloat)
print(numericint / numericfloat)
```

```{r}
## Lists

list <- list(2, 4, 6, 8)
print(list)
```

```{r}
## Booleans

booleant <- TRUE
print(booleant)

booleanf <- FALSE
print(booleanf)
```

```{r}
## Hashes

library(hash)

hashexample <- hash()
hashexample[["key1"]] <- "val1"
hashexample[["key2"]] <- "val2"
hashexample[["key3"]] <- "val3"
print(hashexample)
```

### Variable interactions

Since R is a weakly typed language, it does offer more possibilities for interactions between variables than other languages. For example, integers and floats can interact without being converted. You can also combine variable types in a single list.

```{r}
snlist1 <- list("1", 2, 3)
snlist2 <- list(1, 2.5, 3)

print(snlist1)
print(snlist2)
```

However, R will not allow operations between numbers in string format (for example: "1" as opposed to 1 or 1L)

```{r}
x = "1" + 2
print(2)

# Will throw error: "non-numeric argument to binary operator"
```

### Operations

As a math-based language, it's important to know the artithmetic operators in R:

- Addition: ```+```
- Subtraction: ```-```
- Multiplication: ```*```
- Division: ```/```
- Exponents: ```^```
- Division (returns the remainder): ```%%```
- Integer division: ```%/%```

Notably, assignment operations in R are typically written as ```variable <- value``` or ```value -> variable``` rather than a single equal sign.

There are also logical operators and comparisons:
- Equal-to: ```==```
- Not-equal-to: ```!=```
- Less-than: ```<```
- Greater-than: ```>```
- Less-than or equal-to: ```<=```
- Greater-than or equal-to: ```>=```
- AND (both are TRUE): ```&```
- OR (at least one is TRUE): ```|```

(P.S.: R has secret extra symbols for AND and OR -- see the README section "Short circuiting" for more info)

```{r}
## Arithmetic operations

1+2     # Addition
1-2     # Subtraction
1*2     # Multiplication
1/2     # Division
1^2     # Exponents
1%%2    # Division remainder
1L%/%2L # Integer division
```

```{r}
## Logical operations & comparisons

1+2 == 3  # Equal to
1+2 != 3  # Not equal to

1<2 # Less than
1>2 # Greater than

1<=2  # Less than or equal to
1>=2  # Greater than or equal to
```

```{r}
1 == 1 & 2 == 2   # 1 is equal to 1 AND 2 is equal to 2
1 != 1 & 2 == 2   # 1 is not equal to 1 AND 2 is equal to 2

1 == 1 | 2 == 2   # 1 is equal to 1 OR 2 is equal to 2 (TRUE, TRUE)
1 != 1 | 2 == 2   # 1 is not equal to 1 OR 2 is equal to 2 (FALSE, TRUE)
1 != 1 | 2 != 2   # 1 is not equal to 1 OR 2 is not equal to 2 (FALSE, FALSE)
```

## PLP 3 - Functions

### Declaring a function in R
The syntax to declare a function is ```yourFunctionName <- function(parameters) {definition}```. Functions must be declared before (above) they are called in the code. R does not allow functions to return more than one value, but it does support the ability to pass multiple parameters of different types into a function. It also allows recursive (self-calling) functions.

```{r}
multitype <- function(a, b)
{
  return(paste(a, b))
}

multitype("a", 3)
```

```{r}
factfunct <- function(val)
{
  if(val < 1)
  {
    return(1)
  }
  else
  {
    return(val * factfunct(val-1))
  }
}

factfunct(4)
```

```{r}
factof5 <- factfunct(5)
print(factof5)
```

### Pass-by-reference or pass-by-value?
Whether a language is "pass-by-reference" or "pass-by-value" determines what happens when you assign a variable to another variable. Depending on the language's design, it may copy the value of the old variable into the new one (pass-by-value) or it may create the new variable as an alias of the old variable that will reflect any changes that occur to it (pass-by-reference). We can tell which protocol a language uses by assigning an old variable to a new variable, changing the value of the old variable, and then checking the value of the new variable. If the new value changed with the old variable, the language is pass-by-reference. If the new variable holds the value that the old variable initially started with, it is pass-by-value. As you can see in this example, R is a pass-by-value language.

```{r}
old <- 3

new <- old

old = 4

print(new)
```

### Scope
R is a statically scoped language, meaning that when it encounters "free," unassigned variables, it searches within the smallest possible environment first and then expands its search outward. For example, if the variable <x> is called in function ```function1```, which resides in the function ```main```, R would first search for ```x``` in ```function1```, then ```main```, then through the "search" list. This list always begins with the global environment (the space outside all of your functions), is followed by any packages that you've loaded into your program, and ends with the "base" R code. This order can be modified and may be relevant if there are objects with the same name in different packages.

## PLP 4 - Selection, Loops, and Conditionals

### Boolean values & condition statements
R's boolean values are ```TRUE``` and ```FALSE```. Its conditional statements are ```if```, ```else if```, and ```else```, which are written as follows:

```
# Not actual R code!

if(condition 1){
  action 1
} else if(condition 2) {
  action 2
} else {
  action 3
}
```

The braces and ```else if``` option help to prevent the "dangling else" ambiguity, in which consecutive "if" statements are followed by an "else" and the program must decide which "if" statement it belongs to.

```{r}
trueval <- TRUE
falseval <- FALSE

if(trueval == TRUE){
  print("trueval is true")
} else {
  print("trueval is false")
}

if(falseval == TRUE){
  print("falseval is true")
} else {
  print("falseval is false")
}
```

```{r}
if(trueval == TRUE & falseval == FALSE){
  print("trueval is true and falseval is false")
} else if(falseval == TRUE){
  print("both values are true")
} else if(trueval == FALSE){
  print("both values are false")
} else {
  print("trueval is false and falseval is true")
}
```

### Short circuiting
Short circuiting allows a program to take "shortcuts" in certain cases of logical condition statements. For example, given that ```T``` is TRUE and ```F``` is FALSE:

In the case of ```if(T OR F)```: a program only needs one of the two statements to be TRUE in order to proceed. Since the first item is TRUE, it doesn't matter what the second item is; the condition for the OR statement has been met. If a language uses short circuit logic, the program would skip over the second item and go straight to the designated code block. **This statement short circuits.**

In the case of ```if(F OR T)```: the program needs to see that the second item is TRUE in order for the condition to be met. **This statement does not short circuit.**

In the case of ```if(F AND T)```: the program needs both items to be TRUE in order to proceed. Since it can tell from the first item that this cannot be the case, it will skip over the second item and not execute the designated code block. **This statement short circuits.**

In the case of ```if(T AND F)```: the program needs to see that the second item is FALSE before it can know that the condition is not met. **This statement does not short circuit.**

In R, we get extra versions of the logical operators that indicate AND and OR -- whereas ```&``` and ```|``` do NOT allow short circuiting, the logical operators ```&&``` and ```||``` do! You can see the difference (and the potential consequences) of using one set or the other in the stopIt example in the following examples.

```{r}
stopIt <- function(){
  stop("NOT short circuited!")
}
  
if(1 == 1 || stopIt()){
  print("short circuited!")
}

if(1 == 1 | stopIt()){
  print("short circuited!")
}
```

### Switch
The "switch" function in R allows you to evaluate an expression against any number of "cases" and choose the appropriate path. It takes in an expression, cases, and sometimes an optional default case. 

The first type of switch statement has a string as its expression. Its syntax is as follows:

```
# Not actual R code!

expression <- "string2"

switch(expression,
      "string1" = "case1",
      "string2" = "case2,
      "string3" = "case3,
      "default case") # The default case is optional. If not specified, an unmatched expression will return NULL

# The program will proceed with action2. If the expression was, for example, "string4", or "stringABC", it would proceed with "default case".
```

Interactive example:

```{r}
switchboard <- function(route){
  switch(direction,
         "northwest" = "take track 1",
         "north" = "take track 2",
         "northeast" = "take track 3",
         "this direction is not available")
}

direction <- "north"
print(switchboard(direction))
```

```{r}
direction <- "northeast"
print(switchboard(direction))
```

```{r}
direction <- "east"
print(switchboard(direction))
```

The second type has an integer as its expression. It does not allow for the option of a default case. It will proceed with the case that has the index of the expression integer. Syntax:

```
# Not actual R code!

expression <- 1

switch(expression,
      "case1",
      "case2",
      "case3",
      "case4")

# The program will proceed with "case1" -- remember, R starts counting at 1!
```

Interactive example:

```{r}
switchnum <- function(expression)
  switch(expression,
         "option a",
         "option b",
         "option c",
         "option d")

exp <- 1

print(switchnum(exp))
```

```{r}
exp <- 5

print(switchnum(exp))
```
